---
title: "homework5"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{homework5}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
```

```{r setup}
library(temp)
```

# Question 1

## Part (a)

Here I write the complete log-likelihood for the mixture model.

Let $\mathcal{l}(\alpha, \mu_{1}, \mu_{2})$ denote $\text{ln}\text{Pr}(\bf{x}, \bf{y}\,|\, \alpha, \mu_{1}, \mu_{2})$ 

## Part (b)

Here I obtain an algebraic form of \[E \bigg[ \mathcal{l}(\alpha, \mu_{1}, \mu_{2} \,|\, \bf{y}, \alpha^{k}, \mu_{1}^{k}, \mu_{2}^{k}) \bigg]\], where the expectation is taken with respect to $\text{Pr}(\bf{x}| \bf{y}, \alpha^{(k)}, \mu_{1}^{(k)}, \mu_{2}^{(k)})$

## Part (c) 

Here I maximize the expectation from part (b) to obtain the iterative equations for the EM algorithm


## Part (d)

Here I implement the EM algorithm and run it on the data saved in the file `mixture_data.txt`.


## Part (e)

Here I conduct a simulation study to estimate the bias of the MLE for the mixture model above.











